%% Copyright 2008, The TPIE development team
%% 
%% This file is part of TPIE.
%% 
%% TPIE is free software: you can redistribute it and/or modify it under
%% the terms of the GNU Lesser General Public License as published by the
%% Free Software Foundation, either version 3 of the License, or (at your
%% option) any later version.
%% 
%% TPIE is distributed in the hope that it will be useful, but WITHOUT ANY
%% WARRANTY; without even the implied warranty of MERCHANTABILITY or
%% FITNESS FOR A PARTICULAR PURPOSE.  See the GNU Lesser General Public
%% License for more details.
%% 
%% You should have received a copy of the GNU Lesser General Public License
%% along with TPIE.  If not, see <http:%%www.gnu.org/licenses/>

\documentclass[10pt]{book}
%\usepackage{html} % for inserting html links to be used by latex2html
\ifx\pdfoutput\undefined 
\usepackage[dvips]{hyperref}
\else
\usepackage[pdftex,colorlinks=true]{hyperref}
\fi
\usepackage{path}
\usepackage{makeidx}
\usepackage{verbatim} % new verbatim package.
\usepackage{listings}
%\usepackage{times} % for better pdf output.
\usepackage{color} % for colorbox
\input{definitions}

\usepackage{graphicx}
%% Comment this out to produce distribution version of manual.
\withcomments

\makeindex

\begin{document}

\title{{\Huge TPIE}\\ User Manual and Reference}
\author{%
Lars Arge \and 
Rakesh Barve \and
Andrew Danner \and
David Hutchinson \and 
Thomas M\o lhave, \and
Octavian Procopiuc \and 
J\"{o}rg Rotthowe \and
Laura Toma \and
Jan Vahrenhold \and
Darren Erik Vengroff \and 
Markus Vogel \and
Rajiv Wickeremesinghe}


\date{{\bf DRAFT} of \today}

\maketitle

\begin{titlepage}
\mbox{ }

\vspace{\fill}

\noindent TPIE User Manual and Reference

\noindent Edition \edition, for TPIE version \version.

\vspace{2ex}

\noindent Copyright \copyright 1994, 1995 Darren Erik Vengroff, 2002 Lars
Arge, Rakesh Barve, David Hutchinson, Octavian Procopiuc, Laura Toma,
Darren Erik Vengroff, Rajiv Wickeremesinghe, 2005 Lars Arge, Andrew
Danner, Thomas M\o lhave, Octavian Procopiuc, J\"{o}rg Rotthowe,
Laura Toma, Jan Vahrenhold, Markus Vogel.


\vspace{2ex}

This file is part of TPIE.

TPIE is free software: you can redistribute it and/or modify
it under the terms of the GNU Lesser General Public License as published by
the Free Software Foundation, either version 3 of the License, or
(at your option) any later version.

TPIE is distributed in the hope that it will be useful,
but WITHOUT ANY WARRANTY; without even the implied warranty of
MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
GNU Lesser General Public License for more details.

You should have received a copy of the GNU Lesser General Public License
along with TPIE.  If not, see \texttt{http://www.gnu.org/licenses/}.


\end{titlepage}

\tableofcontents

\chapter*{Introduction}
\addcontentsline{toc}{chapter}{Introduction}

This manual describes TPIE, a Transparent Parallel I/O Environment,
designed to assist programmers in writing high performance
I/O-efficient programs for a variety of platforms.\comment{LA: Before
  distribution add a note about block collection stuff not documented
  yet}

\emph{This manual, like the whole of the TPIE project, is work in
  progress. The authors are making it available in its current state
  in the hopes that it will be useful, but without any warranty
  whatsoever. Refer to the copyright page at the beginning of this
  manual for full details. Please send comments, bug reports\index{bug
    reports}, etc., to \path"tpie@cs.duke.edu".}

\input{acknowledgements} % Chapter: Acknowledgments

\part{User Manual}
  \input{user_manual}

\part{Reference}
  \input{ami_reference} % Chapter: AMI Programmer's Reference
  \input{implementation} % Chapter: The Implementation of TPIE
  \input{tuning} % Chapter: TPIE Performance Tuning
\part{Appendices}
\appendix
  \input{applications} %  Chapter: Test and Sample Applications
  \input{addexamples}
  \input{errorcodes}
  \input{lgpl-3.0}

\bibliographystyle{abbrv}
\newpage
\addcontentsline{toc}{part}{Bibliography}
\bibliography{tpie}
\newpage
\addcontentsline{toc}{part}{Index}
\printindex
\end{document}


